Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-11-15 manuscript.tex
Make PDF in manuscript mode for submission
Remove figure size commands
Add link to software.seg.org in the paper
Clarify what LIFO stands for
Add D values to abstract + extra about the code

\pdfoutput=1
\documentclass[paper,twocolumn,twoside]{geophysics}
%\documentclass[manuscript]{geophysics}
%\documentclass[manuscript,endfloat]{geophysics}
%\documentclass[twocolumn]{article}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\newcommand{\mitbf}[1]{\mathbf{{#1}}}

@@

\righthead{Tesseroids: modeling in spherical coordinates}
\ms{Submission} % manuscript number

\address{
\footnotemark[1]Universidade do Estado do Rio de Janeiro,
Rio de Janeiro, Brazil.
email: leouieda@gmail.com
\\
\footnotemark[2]Observat\'orio Nacional,
Rio de Janeiro, Brazil.
email: valcris@on.br
\\
\footnotemark[3]Department of Mathematics and Geosciences,
University of Trieste, Trieste, Italy
}

\begin{abstract}
We present the open-source software \emph{Tesseroids},
a set of command-line programs to perform the forward modeling
of gravitational fields in spherical coordinates.
The fields are calculated numerically using
the Gauss-Legendre Quadrature (GLQ).
We have improved upon an adaptive discretization algorithm
to guarantee the accuracy of the GLQ integration.
Our implementation of adaptive discretization
uses a  ``stack'' based algorithm
instead of recursion to achieve higher performance
and more control over execution errors and corner cases.
The algorithm is controlled by
a scalar called the distance-size ratio ($D$)
that determines the accuracy of the integration as well as
the computation time.
We determined optimal values of $D$
for the gravitational potential, gravitational acceleration,
and gravity gradient tensor
by comparing the computed tesseroids effects
with those of a homogeneous spherical shell.
Contrary to previous assumptions,
our results show that the potential and its first and second derivatives
require different values of $D$ to achieve a target accuracy of 0.1\%.
These values were incorporated as defaults in the software.
\end{abstract}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/tesseroid}
    \caption{
        View of a tesseroid,
        the integration point $Q$ inside the tesseroid,
        a geocentric coordinate system $(X, Y, Z)$,
        the computation $P$ and it's local coordinate system $(x, y, z)$.
        $r$, $\phi$, $\lambda$ are
        the radius, latitude, and longitude, respectively, of point $P$,
        and $\ell$ is the Cartesian distance between $P$ and $Q$.
    }
    \label{fig:tesseroid}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:division}
\end{figure}

Instead of using recursive function calls,
as originally proposed by \citet{Li2011},
we use a stack-based implementation of the algorithm.
Stacks are array-like data structures
with a particular way of inserting and removing elements from it.
In a stack,
one can only insert elements to the top of the stack
(the last empty position).
Likewise,
one can only remove the last element of the stack
(commonly referred to as ``popping'' the stack).
Stacks are also known as ``Last-in-first-out'', or LIFO, data structures.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/distance-size-curves}
    \caption{
        The maximum difference between the computed tesseroid and shell effects
        as a function of the distance-size ratio $D$
        for (a) the gravitational potential, (b) $g_z$, and (c) $g_{zz}$.
        The difference is given as a percentage of the shell effect.
        Curves correspond to the different tesseroid models and computation
        grids shown in Table~\ref{tbl:experiment}.
        The horizontal solid black line marks the established error threshold
        of $0.1\%$.
        A value of $D=0$ means that no divisions are made.
    }
    \label{fig:dist-size-curves}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/gzz-with-height}
    \caption{
        Difference between the computed $g_{zz}$ for the spherical shell and
        the tesseroid model at different heights. Curves show the maximum
        difference as a percentage of the shell value.
        The horizontal solid black line marks the established error threshold
        of $0.1\%$.
        A value of $D=0$ means that no divisions are made.
    }
    \label{fig:gzz-with-height}
\end{figure}

>>>>

\pdfoutput=1
%\documentclass[paper,twocolumn,twoside]{geophysics}
%\documentclass[manuscript]{geophysics}
\documentclass[manuscript,endfloat]{geophysics}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\newcommand{\mitbf}[1]{\mathbf{{#1}}}

@@

\righthead{Tesseroids: modeling in spherical coordinates}
\ms{GEO-2015-0204} % manuscript number

\address{Peer-reviewed code related to this article can be found at
http://software.seg.org/2016/000X.
\\
\footnotemark[1]Universidade do Estado do Rio de Janeiro,
Rio de Janeiro, Brazil.
email: leouieda@gmail.com
\\
\footnotemark[2]Observat\'orio Nacional,
Rio de Janeiro, Brazil.
email: valcris@on.br
\\
\footnotemark[3]Department of Mathematics and Geosciences,
University of Trieste, Trieste, Italy
}

\begin{abstract}
We present the open-source software \emph{Tesseroids},
a set of command-line programs to perform the forward modeling
of gravitational fields in spherical coordinates.
The software is implemented in the C programming language and uses tesseroids
(spherical prisms) for the discretization of the subsurface mass distribution.
The gravitational fields of tesseroids are calculated numerically using
the Gauss-Legendre Quadrature (GLQ).
We have improved upon an adaptive discretization algorithm
to guarantee the accuracy of the GLQ integration.
Our implementation of adaptive discretization
uses a ``stack'' based algorithm
instead of recursion to achieve
more control over execution errors and corner cases.
The algorithm is controlled by
a scalar value called the distance-size ratio ($D$)
that determines the accuracy of the integration as well as
the computation time.
We determined optimal values of $D$
for the gravitational potential, gravitational acceleration,
and gravity gradient tensor
by comparing the computed tesseroids effects
with those of a homogeneous spherical shell.
The values required for a maximum relative error of 0.1\% of the shell effects
are $D = 1$ for the gravitational potential, $D = 1.5$ for the gravitational
acceleration, and $D = 8$ for the gravity gradients.
Contrary to previous assumptions,
our results show that the potential and its first and second derivatives
require different values of $D$ to achieve the same accuracy.
These values were incorporated as defaults in the software.
\end{abstract}

@@

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid}
    \caption{
        View of a tesseroid,
        the integration point $Q$ inside the tesseroid,
        a geocentric coordinate system $(X, Y, Z)$,
        the computation $P$ and it's local coordinate system $(x, y, z)$.
        $r$, $\phi$, $\lambda$ are
        the radius, latitude, and longitude, respectively, of point $P$,
        and $\ell$ is the Cartesian distance between $P$ and $Q$.
    }
    \label{fig:tesseroid}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:division}
\end{figure}

Instead of using recursive function calls,
as originally proposed by \citet{Li2011},
we use a stack-based implementation of the algorithm.
Stacks are array-like data structures
with a particular way of inserting and removing elements from it.
In a stack,
one can only insert elements to the top of the stack
(the last empty position).
Likewise,
one can only remove the last element of the stack
(commonly referred to as ``popping'' the stack).
Because of these restrictions,
stacks are also known as ``Last-In-First-Out'' (LIFO) data structures.

@@

\begin{figure}
    \centering
    \includegraphics{figs/distance-size-curves}
    \caption{
        The maximum difference between the computed tesseroid and shell effects
        as a function of the distance-size ratio $D$
        for (a) the gravitational potential, (b) $g_z$, and (c) $g_{zz}$.
        The difference is given as a percentage of the shell effect.
        Curves correspond to the different tesseroid models and computation
        grids shown in Table~\ref{tbl:experiment}.
        The horizontal solid black line marks the established error threshold
        of $0.1\%$.
        A value of $D=0$ means that no divisions are made.
    }
    \label{fig:dist-size-curves}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{figs/gzz-with-height}
    \caption{
        Difference between the computed $g_{zz}$ for the spherical shell and
        the tesseroid model at different heights. Curves show the maximum
        difference as a percentage of the shell value.
        The horizontal solid black line marks the established error threshold
        of $0.1\%$.
        A value of $D=0$ means that no divisions are made.
    }
    \label{fig:gzz-with-height}
\end{figure}
